JD notes on effects

### 2022-07-20 manuscript/body.tex
share with Bicko

We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and observed-value-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias. In addition, we demonstrate that observed-value-based approach can be used for bias correction.

@@

The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively. However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, averaging the interactions (averaging the product of linear predictor variables) versus taking the product of the averages of the input variables (the default for \pkg{emmeans} and \pkg{effects}). For simple OLS models (with no interactions), the two approaches give the same estimates. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values. To illustrate this, consider models description by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} below, with $x_1$ as the focal predictor:
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
\end{align}
%
We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (based on the true simulation parameters) and observed average, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.inter.pdf}
\end{center}
\caption{{\bf A comparison of central estimates for \pkg{emmeans}, \pkg{effects} and \pkg{varpred}.} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are the binned simulated $y$. The trend lines represent the $\hat{y}$ -- central estimate. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated one in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the central estimate based on the three approaches match the truth. B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean, $\bar{y}$. Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}
%
In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A. However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B. To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the linear predictor variables corresponding to $x_2$ and $x_3$. On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction linear predictor variable ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses linear predictor variables to compute $\over{x_2x_3}$. In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the linear predictor variables.

>>>>

We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and observed-value-based approaches produce different results in models with nonlinear link functions \jd{What‽: and/or additional source(s) of potential bias}. In addition, we demonstrate that observed-value-based approach can be used for bias correction.

@@

The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively. However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, the default for \pkg{emmeans} and \pkg{effects} is to average the predictor variables and use these averaged values for the interaction, as opposed to our preferred model-center approach (averaging each linear predictor variable separately). The packages give the same results as our method for models without interactions, but when interactions are present, the two approaches can produce substantially different results, with the model-center approach more closely matching the observed values.

To illustrate this, we simulate data from the models below:
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
\end{align}
%
We simulate using $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\{x_{1,2,3}, \epsilon\} \sim \mathrm{Normal}(0, 1)$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (calculated using the simulation parameters) and observed average, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.inter.pdf}
\end{center}
\caption{{\bf A comparison of central estimates for \pkg{emmeans}, \pkg{effects} and \pkg{varpred}.} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are averages of the simulated points binned according to value of $x_1$. The trend lines represent the $\hat{y}$ -- central estimate. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, is the same in all three approaches, and closely matches the true mean; the central estimate likewise matches the truth (horizontal lines). B: With a simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean, $\bar{y}$. Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}
%
In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A. However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B. To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the linear predictor variables corresponding to $x_2$ and $x_3$. On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction linear predictor variable ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses linear predictor variables to compute $\over{x_2x_3}$. In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the linear predictor variables.
